---
title: "Assignment 1"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**NAME: Nicholas Tolley**\
**DUE DATE: February 14th, 6pm**

Please post your completed homework (including the R code).

## Problem I (100 pts)

This exercise focuses on the data sets country.rda, which can be found in Canvas in the Files item in the folder Data (file name: country.rda). It contains data called gapminder from the Gapminder non-profit organization. The data are 10,545 units and 9 variables. The variables are:

country

continent

year

infant_mortality: infant mortality rate (to interpret, if you divide by 10 you will obtain the percentage)

life_expectancy

fertility

population

gdp

continent

region

Read and load the data in R.

```{r}
library(tidyverse)
library(dslabs)
library(dplyr)

load(file='country.rda')
```

1)  Perform a boxplot with ggplot in R where you can see in the x axis each continent and y axis the population just for year 1960, hint: use population/10\^6 and use the log transformation. Be careful on writing the axes labels. What continent has the largest median population size? Check in R. Explain what you have found with theoretical details (outliers, symmetry of the distribution, median and interquartile range).

```{r}
filtered_data_1960 <- filter(gapminder, year==1960)
p <- ggplot(filtered_data_1960, aes(x=continent, y=population/10^6, fill=continent)) + geom_boxplot() + scale_y_log10() + theme_minimal() + labs(y="log population (in millions)", title="Continental population size in 1960")
p
```

Based on the plot above, the continent with the largest median population size is Europe, meaning that countries in Europe had generally larger populations than other continents in 1960. The code below calculates the exact median population size:

```{r}
filter(filtered_data_1960, continent=="Europe") %>%  pull(population) %>% median()

```

As we can see, the median population size of Europe in 1960 was 7,065,525. In general we see that populations for each continent are symmetric around the median, that is the 1st and 3rd quartiles are roughly equidistant from the median. The notable exception being Oceania with a clear positive skew. However, noting that this data is a plotted on a log transformed y-axis, we can conclude that the country specific populations for all continents exhibit a rightward skew.

Only two continents exhibit outliers, with Africa having a country with a country population much smaller than the others. In contrast we see that Europe has outlier countries that are both larger and smaller than 1.5x the IQR (the threshold for being an outlier).

2)  Perform a boxplot with ggplot where you can see in the x axis each continent and y axis the population just for year 2010, hint: use population/10\^6 and use the log transformation. What continent has the largest median population size in this year? Check in R. Explain what you have found with theoretical details (outliers, symmetry of the distribution, median and interquartile range).

```{r}
filtered_data_2010 <- filter(gapminder, year==2010)
p <- ggplot(filtered_data_2010, aes(x=continent, y=population/10^6, fill=continent)) + geom_boxplot() + scale_y_log10() + theme_minimal() + labs(y="log population (in millions)", title="Continental population size in 2010")
p
```

3)  What is the median population size for Africa to the nearest million? What proportion of countries in Europe have populations below 14 million? Perform in R.

The code below calculates the median population size for Africa

```{r}
filter(gapminder, continent=="Africa") %>%  pull(population) %>% median(na.rm=TRUE) / 10^6

```

The code below calculates the proportion of countries in Europe with populations below 14 million.

```{r}
filtered_data_europe <- filter(gapminder, continent=="Europe")
num_countries <- nrow(filtered_data_europe)
nrow(filter(filtered_data_europe, population < (14 * 10^6))) / num_countries
```

4)  If we use a log transformation, which continent has the largest interquartile range? Perform and plot in R. Explain what you have found, give also the theoretical implications.

As the plot below shows, across all years in the data set the Americas show the largest interquartile range on log transformed data.

```{r}
iqr_table <- group_by(gapminder, continent) %>% summarize(IQR=IQR(log10(population), na.rm=TRUE))
p <- ggplot(iqr_table, aes(x=continent, y=IQR, fill=continent)) + geom_col() + theme_minimal() + labs(y='log(IQR of population)')
p

```

5)  Compare (in ggplot) the countries in term of population using boxplots with two different colours for year 1960 and 2010. Try to make a plot exactly as Figure 1 (just the colour can change).

```{r}
compare_df <- rbind(filtered_data_1960, filtered_data_2010)
compare_df$year <- as.character(compare_df$year)
p <- ggplot(compare_df, aes(x=continent, y=population / (10^6), fill=year)) + geom_boxplot() + theme_bw() + scale_y_log10() + labs(y='Population in millions')
p

```

```{r, out.width="0.8\\linewidth", include=TRUE, fig.align="center", fig.cap=c("Population per country"), echo=FALSE}
knitr::include_graphics("plot1.pdf")
```

6)  Compare (in ggplot) the countries in term of infant mortality using boxplots with two different colours for year 1960 and 2010. Try to make a plot exactly as Figure 2 (just the colour can change).

```{r}
p <- ggplot(compare_df, aes(x=continent, y=infant_mortality, fill=year)) + geom_boxplot(na.rm = TRUE) + theme_bw() + labs(y='Infant Mortality') + scale_y_log10()
p
```

```{r, out.width="0.8\\linewidth", include=TRUE, fig.align="center", fig.cap=c("Infant Mortality per country"), echo=FALSE}
knitr::include_graphics("plot2.pdf")
```

7)  What proportion of the infant mortality is between 100 and 150? And what is its approximation to a normal distribution of this proportion? What proportion of the infant mortality is greater than 150? And what is its approximation to a normal distribution of the proportions? Perform in R. Explain what you have found, give also the theoretical implications and what theorem or statistical assumptions you are using.

The code below calculates the proportion of infant mortality between 100 and 150 across all years/countries in the dataset. We can see that only 12.6% of the dataset fits this requirement.

```{r}
infant_mortality_100_150 <- filter(gapminder, infant_mortality <= 150, infant_mortality > 100)
nrow(infant_mortality_100_150) / nrow(gapminder) 

```

To find the approximation to the normal distribution of this proportion, we simply need to calculate the mean and standard deviation of the data. This parameterizes a normal approximation, which can be visualized against a histogram of the real data. To apply the normal approximation we are assuming that every entry for infant mortality in the gapminder dataset is independent and identically distributed (IID). Unfortunately we cannot make the first assumption of independence as we already know that the various entries in the dataset span multiple years for the same country.

Additionally, from the previous question we know that the countries belonging to different continents have infant mortality rates with a mean that changes dramatically across years. This means that we can't assume the data is identically distributed, meaning that the assumptions of the normal distribution are not met. Nevertheless, we can still visualize the theoretical normal on top of the empirical histogram.

```{r}
mean_100_150 <- mean(infant_mortality_100_150$infant_mortality, na.rm = TRUE)
sd_100_150 <- sd(infant_mortality_100_150$infant_mortality, na.rm = TRUE)

p <- ggplot(infant_mortality_100_150, aes(x=infant_mortality)) + geom_histogram(aes(y = ..density..),fill='#56B4E9', na.rm = TRUE) + stat_function(fun = dnorm, args = list(mean = mean_100_150, sd = sd_100_150), size=2) + xlim(75, 175)
p
```

As we can see the normal distribution is a poor approximation for this data as the tails are bounded at 100 and 150, and the mean is clearly different from the mode of the empirical data. A uniform distribution may be more appropriate here.

Repeating the same analysis as above, we see that only 3.9% of the dataset has an infant mortality above 150. Additionally we see that normal distribution is still not a good approximation for this data which exhibits a strong rightward skew, making the mean of the empirical data (and therefore the normal distribution) far from the empirical histogram.

```{r}
infant_mortality_150 <- filter(gapminder, infant_mortality > 150)
nrow(infant_mortality_150) / nrow(gapminder) 
```

```{r}
mean_150 <- mean(infant_mortality_150$infant_mortality, na.rm = TRUE)
sd_150 <- sd(infant_mortality_150$infant_mortality, na.rm = TRUE)

p <- ggplot(infant_mortality_150, aes(x=infant_mortality)) + geom_histogram(aes(y = ..density..),fill='#56B4E9', na.rm = TRUE) + stat_function(fun = dnorm, args = list(mean = mean_150, sd = sd_150), size=2) + xlim(145, 300)
p

```

8) In ggplot perform both a histogram and a curve for infant mortality, use two different colours for year 1960 and 2010. Does the infertility follow a normal distribution? Are the two distribution symmetric or skewed? Be careful to choose the right x axis for the histogram.

Neither of the infant mortality rates for 1960 or 2010 follow a normal distribution indicated by the poor agreement between the empirical histogram and the normal approximation. Instead we see that the distributions exhibit a rightward skew, with the mode of the data being smaller than the mean/median. We can see this by overlaying the normal approximation of both distributions whose center indicates the true mean of the data.

```{r}

p <- ggplot(compare_df, aes(x=infant_mortality, fill=year)) + geom_histogram(aes(y = ..density..), na.rm = TRUE) + stat_function(fun = dnorm, args = list(mean = mean(filtered_data_1960$infant_mortality, na.rm = TRUE), sd = sd(filtered_data_1960$infant_mortality, na.rm = TRUE)), size=2) + stat_function(fun = dnorm, args = list(mean = mean(filtered_data_2010$infant_mortality, na.rm = TRUE), sd = sd(filtered_data_2010$infant_mortality, na.rm = TRUE)), size=2) + theme_minimal()

```

9)  Perform a q-q plot for the infant mortality to check if this variables is normal or not. Write the label of the axes in the appropriate way. Explain what you have found including the theoretical implications.

The code below calculates the QQ plot for infant mortality. As we can see, the points representing do not lie along the diagonal line which represents perfect agreement between the theoretical (x axis) and empirical quantiles (y axis). The conclusion is that the data for infant mortality are not normally distributed in the gapminder dataset.

Instead we see that lowest and largest values of observed infant mortality rate are larger than those indicates by the theoretical quantile if the data were normally distributed. We can interpret this as the data having a slight rightward skew.

```{r}
p <- seq(0.05, 0.95, 0.05)

sample_quantiles <- quantile(gapminder$infant_mortality, p, na.rm=TRUE)
theoretical_quantiles <- qnorm(p, mean = mean(gapminder$infant_mortality,na.rm=TRUE), sd = sd(gapminder$infant_mortality, na.rm=TRUE))

qplot(theoretical_quantiles, sample_quantiles) + geom_abline()
```

10) Compare the regions in term of infant mortality for the two years 1960 and 2010 using the barplot. Try to make a plot exactly as Figure 3 (just the colour can change). Create a binary variable, with value 1 if life-expectancy is greater than 65 (group A), 0 otherwise (group B). Then perform the appropriate two sample test to know if the mean of the infant mortality is the same in group A and B (remember to check the variance of the two groups). Explain your results.

```{r}
max_infant_mortality <- summarize(group_by(compare_df, region, year), infant_mortality=max(infant_mortality, na.rm=TRUE))
max_infant_mortality$infant_mortality[is.infinite(max_infant_mortality$infant_mortality)] <- NA
                                  
                                  
p <- ggplot(max_infant_mortality, aes(x=infant_mortality, y=region, fill=year, group=region)) + geom_col(position='dodge2', na.rm=TRUE) + theme_bw() + labs(x='Infant Mortality', y='Region')
p
```

```{r, out.width="0.8\\linewidth", include=TRUE, fig.align="center", fig.cap=c("Infant Mortality per region"), echo=FALSE}
knitr::include_graphics("plot3.pdf")
```

The code below separates the infant mortality data for all entries in the gapminder dataset into groups A and B based on life expectancy being above or below 65, and subsequently calculates the variance of each group.

```{r}
life_expectancy_mask <- (gapminder['life_expectancy'] > 65)[, 1]
group_A <- filter(gapminder, life_expectancy_mask)
group_B <- filter(gapminder, !life_expectancy_mask)

print('Group A var')
var(group_A$infant_mortality, na.rm = TRUE)
print('Group B var')
var(group_B$infant_mortality, na.rm = TRUE)


```

As we can see, the variance of infant mortality in group B is substantially higher than group A. As a result we should not use the generic t test and should explicitly specify that the variance between groups are not equal for the two sample t test.

```{r}
t.test(group_A$infant_mortality, group_B$infant_mortality, var.equal=FALSE, conf.level=0.95)

```

Running the t test above, we can see that mean infant mortalities of groups A and B are significantly different from one another. What this indicates is that across all countries/years in the gapminder dataset, countries with a life expectancy greater than 65 on average have lower infant mortality rates. Specifically a mean infant mortality rate of 22.9 for group A vs. 97.7 for group B.
